Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2024-02-08 main.tex
:pencil: Add parameters

	\emph{Notation:}
	Italic, bold lower-case, and bold upper-case letters indicate scalars, vectors and matrices, respectively.
	$\jmath$ denotes the imaginary unit.
	$\mathbb{C}$ represents the set of complex numbers.
	$\mathbb{U}^{n \times n}$ denotes the set of $n \times n$ unitary matrices.
	$\mathbf{0}$ and $\mathbf{I}$ are the all-zero and identity matrices with appropriate size, respectively.
	$\Re\{\cdot\}$ takes the real part of a complex number.
	$\mathrm{tr}(\cdot)$ and $\det(\cdot)$ evaluates the trace and determinant of a square matrix, respectively.
	$\mathrm{diag}(\cdot)$ constructs a square matrix with arguments on the main diagonal and zeros elsewhere.
	$\mathrm{sv}(\cdot)$ returns the singular value vector.
	$\sigma_n(\cdot)$ and $\lambda_n(\cdot)$ is the $n$-th largest singular value and eigenvalue, respectively.
	% $\boldsymbol{\sigma}(\cdot)$ and $\boldsymbol{\lambda}(\cdot)$ are the corresponding vectors.
	$(\cdot)^*$, $(\cdot)^\mathsf{T}$, $(\cdot)^\mathsf{H}$, $(\cdot)^{(r)}$, $(\cdot)^{\star}$ denote the conjugate, transpose, conjugate transpose (Hermitian), $r$-th iterated point, and final solution, respectively.
	$(\cdot)_{[x:y]}$ is a shortcut for $(\cdot)_x,(\cdot)_{x+1},\ldots,(\cdot)_y$.
	$\lvert \cdot \rvert$ denotes the absolute value.
	% $\lVert \cdot \rVert _p$ means the $p$-norm and $\lVert \cdot \rVert$ suggests $p = 2$.
	$\lVert \cdot \rVert$ means the Euclidean norm.
	$\lVert \cdot \rVert _\mathrm{F}$ represents the Frobenius norm.
	$\mathcal{CN}(\mathbf{0}, \mathbf{\Sigma})$ is the multivariate \gls{cscg} distribution with mean $\mathbf{0}$ and covariance $\mathbf{\Sigma}$.
	$\sim$ means ``distributed as''.
\end{section}

@@

\begin{section}{Channel Singular Value Redistribution}
	\begin{subsection}{A Toy Example}
		We first illustrate the channel shaping capabilities of different \gls{ris} by a toy example.
		% We first use a toy example to illustrate that \gls{bd}-\gls{ris} can provide a wider dynamic range of channel singular values.
		Consider a $2 \times 2 \times 2$ setup where the direct link is blocked.
		The diagonal \gls{ris} is modeled by $\mathbf{\Theta}_\mathrm{D} = \mathrm{diag}(e^{\jmath \theta_1}, e^{\jmath \theta_2})$ while the unitary \gls{bd}-\gls{ris} has 4 independent angular parameters
		\begin{equation}
			\mathbf{\Theta}_\mathrm{U} = e^{\jmath \phi} \begin{bmatrix}
				e^{\jmath \alpha} \cos \psi  & e^{\jmath \beta} \sin \psi   \\
				-e^{-\jmath \beta} \sin \psi & e^{-\jmath \alpha} \cos \psi
			\end{bmatrix}.
		\end{equation}
		In particular, $\phi$ has no impact on the singular value because $\mathrm{sv}(e^{\jmath \phi} \mathbf{A}) = \mathrm{sv}(\mathbf{A})$.
		We also enforce symmetry by $\beta = \pi / 2$ such that both architectures have the same number of angular parameters.
		\begin{figure}
			\centering
			\includegraphics[width=\columnwidth]{assets/simulation/pc_singular_toy.eps}
			\caption{$2 \times 2 \times 2$ channel singular value shaping by diagonal and symmetry unitary \gls{ris}. Direct link is absent.}
			\label{sm:pc_singular_toy}
		\end{figure}
		Fig.~\ref{sm:pc_singular_toy} shows the channel singular values achieved by an exhaustive grid search over $(\theta_1, \theta_2)$ for diagonal \gls{ris} and $(\alpha, \psi)$ for symmetric unitary \gls{ris}.
		It is observed that both singular values can be manipulated up to $9\%$ using diagonal \gls{ris} and $42\%$ using symmetric \gls{bd}-\gls{ris}, despite both architectures have the same number of scattering elements and design parameters.
		A larger performance gap is expected when asymmetric \gls{bd}-\gls{ris} is available.
		This example shows \gls{bd}-\gls{ris} can provide a wider dynamic range of channel singular values and motivates further studies on channel shaping.
	\end{subsection}

@@

		% The Pareto singular value problem \eqref{op:pareto} can thus be solved by Algorithm~\ref{ag:rcg}.
		\begin{lemma}
			The Euclidean gradient of \eqref{ob:pareto} w.r.t. \gls{bd}-\gls{ris} group $g$ is
			\begin{equation}
				\frac{\partial \sum_n \rho_n \sigma_n(\mathbf{H})}{\partial \mathbf{\Theta}_g^*} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \mathbf{U} \mathrm{diag}(\rho_1,\ldots,\rho_N) \mathbf{V}^\mathsf{H} \mathbf{H}_{\mathrm{F},g}^\mathsf{H},
				\label{eq:pareto_gradient}
			\end{equation}
			where $\mathbf{U} \in \mathbb{C}^{N_\mathrm{R} \times N}$ and $\mathbf{V} \in \mathbb{C}^{N_\mathrm{T} \times N}$ are the left and right compact singular matrices of $\mathbf{H}$, respectively.
		\end{lemma}
		\begin{proof}
			Let $\mathbf{H} = \sum_n \mathbf{u}_n \sigma_n \mathbf{v}_n^\mathsf{H}$ be the compact \gls{svd} of the equivalent channel.
			Since the singular vectors are orthonormal, the $n$-th singular value can be expressed as
			\begin{equation}
				\sigma_n = \mathbf{u}_n^\mathsf{H} \mathbf{H} \mathbf{v}_n = \mathbf{u}_n^\mathsf{T} \mathbf{H}^* \mathbf{v}_n^*,
			\end{equation}
			whose differential w.r.t. $\mathbf{\Theta}_g^*$ is
			\begin{align*}
				\partial \sigma_n
				& = \partial \mathbf{u}_n^\mathsf{T} \underbrace{\mathbf{H}^* \mathbf{v}_n^*}_{\sum_m \mathbf{u}_m^* \sigma_m \mathbf{v}_m^\mathsf{T} \mathbf{v}_n} + \mathbf{u}_n^\mathsf{T} \cdot \partial \mathbf{H}^* \cdot \mathbf{v}_n^* + \underbrace{\mathbf{u}_n^\mathsf{T} \mathbf{H}^*}_{\mathbf{u}_n^\mathsf{T} \sum_m \mathbf{u}_m^* \sigma_m \mathbf{v}_m^\mathsf{T}} \partial \mathbf{v}_n^*\\
				& = \underbrace{\partial \mathbf{u}_n^\mathsf{T} \mathbf{u}_n^*}_{\partial 1 = 0} \cdot \sigma_n + \mathbf{u}_n^\mathsf{T} \cdot \partial \mathbf{H}^* \cdot \mathbf{v}_n^* + \sigma_n \cdot \underbrace{\mathbf{v}_n^\mathsf{T} \partial \mathbf{v}_n^*}_{\partial 1 = 0}\\
				& = \mathbf{u}_n^\mathsf{T} \mathbf{H}_{\mathrm{B},g}^* \cdot \partial \mathbf{\Theta}_g^* \cdot \mathbf{H}_{\mathrm{F},g}^* \mathbf{v}_n^*\\
				& = \mathrm{tr}(\mathbf{H}_{\mathrm{F},g}^* \mathbf{v}_n^*\mathbf{u}_n^\mathsf{T} \mathbf{H}_{\mathrm{B},g}^* \cdot \partial \mathbf{\Theta}_g^*).
			\end{align*}
			According to \cite{Hjorungnes2007}, the corresponding complex derivative is
			\begin{equation}
				\frac{\partial \sigma_n}{\partial \mathbf{\Theta}_g^*} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \mathbf{u}_n \mathbf{v}_n^\mathsf{H} \mathbf{H}_{\mathrm{F},g}^\mathsf{H}.
				\label{eq:derivative_sv}
			\end{equation}
			Linear combination of \eqref{eq:derivative_sv} yields \eqref{eq:pareto_gradient}.
		\end{proof}

@@

	\begin{subsection}{Some Analytical Bounds}
		We then discuss some analytical bounds related to channel singular values.
		\begin{proposition}[rank-deficient channel]
			In point-to-point \gls{mimo}, \gls{bd}-\gls{ris} cannot achieve a higher \gls{dof} than diagonal \gls{ris}.
		\end{proposition}
		\begin{proof}
			The scattering matrix of \gls{bd}-\gls{ris} can be decomposed as\footnote{This is because (block) unitary matrices are closed under multiplication.}
			\begin{equation}
				\mathbf{\Theta} = \mathbf{L} \mathbf{\Theta}_\mathrm{D} \mathbf{R}^\mathsf{H},
			\end{equation}
			where $\mathbf{\Theta}_\mathrm{D} \in \mathbb{U}^{N_\mathrm{S} \times N_\mathrm{S}}$ corresponds to diagonal \gls{ris} and $\mathbf{L}, \mathbf{R} \in \mathbb{U}^{N_\mathrm{S} \times N_\mathrm{S}}$ are block diagonal matrices of $L \times L$ unitary blocks.
			Manipulating $\mathbf{L}$ and $\mathbf{R}$ rotates the linear spans of $\bar{\mathbf{H}}_\mathrm{B} \triangleq \mathbf{H}_\mathrm{B} \mathbf{L}$ and $\bar{\mathbf{H}}_\mathrm{F} \triangleq \mathbf{R}^\mathsf{H} \mathbf{H}_\mathrm{F}$ and maintains their rank.
			On the other hand, there exists a $\mathbf{\Theta}_\mathrm{D}$ such that
			\begin{equation*}
				\begin{split}
					\mathrm{rank}(\mathbf{H}_\mathrm{B} \mathbf{\Theta}_\mathrm{D} \mathbf{H}_\mathrm{F})
					& = \min \bigl( \mathrm{rank}(\mathbf{H}_\mathrm{B}), \mathrm{rank}(\mathbf{\Theta}_\mathrm{D}), \mathrm{rank}(\mathbf{H}_\mathrm{F}) \bigr) \\
					& = \min \bigl( \mathrm{rank}(\bar{\mathbf{H}}_\mathrm{B}), N_\mathrm{S}, \mathrm{rank}(\bar{\mathbf{H}}_\mathrm{F}) \bigr) \\
					& = \max_\mathbf{\Theta} \ \mathrm{rank}(\mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F})
				\end{split}
			\end{equation*}
			The same result holds if the direct link is present.
		\end{proof}

		\begin{proposition}[\gls{los} forward\footnote{A similar result holds for \gls{los} backward channel.} channel]
			If the forward channel is rank-1, then \gls{bd}-\gls{ris} can at most enlarge (resp. suppress) the $n$-th ($n \ge 2$) channel singular value to the $(n-1)$-th (resp. $n$-th) singular value of $\mathbf{T}$, that is,
			\begin{equation}
				\sigma_1(\mathbf{T}) \ge {\sigma_2(\mathbf{H})} \ge \sigma_2(\mathbf{T}) \ge \ldots \ge \sigma_{N-1}(\mathbf{T}) \ge {\sigma_N(\mathbf{H})} \ge \sigma_N(\mathbf{T}),
				\label{iq:sv_bound_los}
			\end{equation}
			where $\mathbf{T} \mathbf{T}^\mathsf{H} = \mathbf{H}_\mathrm{D} (\mathbf{I} - \mathbf{v}_\mathrm{F} \mathbf{v}_\mathrm{F}^\mathsf{H}) \mathbf{H}_\mathrm{D}$ and $\mathbf{v}_\mathrm{F}$ is the right compact singular vector of $\mathbf{H}_\mathrm{F}$.
			Note that $\sigma_1(\mathbf{H})$ is unbounded with a sufficiently large \gls{ris}.
		\end{proposition}
		\begin{proof}
			Let $\mathbf{H}_\mathrm{F} = \sigma_\mathrm{F} \mathbf{u}_\mathrm{F} \mathbf{v}_\mathrm{F}^\mathsf{H}$ be the compact \gls{svd} of the forward channel.
			The channel Gram matrix can be written as Hermitian-plus-rank-1:
			\begin{equation}
				\mathbf{G} \triangleq \mathbf{H} \mathbf{H}^\mathsf{H} = \mathbf{Y} + \mathbf{z} \mathbf{z}^\mathsf{H},
			\end{equation}
			where $\mathbf{Y} \triangleq \mathbf{H}_\mathrm{D}^\mathsf{H} (\mathbf{I} - \mathbf{v}_\mathrm{F} \mathbf{v}_\mathrm{F}^\mathsf{H}) \mathbf{H}_\mathrm{D}$ and $\mathbf{z} \triangleq \sigma_\mathrm{F} \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{u}_\mathrm{F} + \mathbf{H}_\mathrm{D} \mathbf{v}_\mathrm{F}$.
			By the Cauchy interlacing formula \cite{Golub2013}, the $n$-th ($n \ge 2$) eigenvalues of $\mathbf{G}$ are bounded by
			\begin{equation}
				\lambda_1(\mathbf{Y}) \ge \lambda_2(\mathbf{G}) \ge \lambda_2(\mathbf{Y}) \ge \ldots \ge \lambda_{N-1}(\mathbf{Y}) \ge \lambda_N(\mathbf{G}) \ge \lambda_N(\mathbf{Y}).
				\label{iq:ev_bound_los}
			\end{equation}
			Since $\mathbf{Y} = \mathbf{T} \mathbf{T}^\mathsf{H}$ is positive semi-definite, the eigenvalues are non-negative and taking the square root of \eqref{iq:ev_bound_los} gives \eqref{iq:sv_bound_los}.
		\end{proof}
		It is worth notice that a similar conclusion holds for diagonal \gls{ris} \cite{Semmler2023}.
		We will later show that for a finite $N_\mathrm{S}$, using a larger group size can approach those bounds better.

@@

		\begin{remark}
			Interestingly, \eqref{iq:sv_bound_fc}--\eqref{iq:sv_bound_fc_power} are simultaneously tight when $\mathbf{\Theta} = \mathbf{V}_\mathrm{B} \mathbf{U}_\mathrm{F}^\mathsf{H}$.
			From \eqref{eq:scattering_fc} and \eqref{eq:channel_equivalent_fc}, we conclude the off-diagonal entries can enhance the capabilities of
			\begin{itemize}
				\item subspace alignment: $\mathbf{V}_\mathrm{B}$ and $\mathbf{U}_\mathrm{F}^\mathsf{H}$ in \eqref{eq:scattering_fc} fully align the subspaces of $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ by rotation;
				\item subchannel rearrangement: $\mathbf{X} = \mathbf{I}$ in \eqref{eq:channel_equivalent_fc} pairs the subchannels of $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ from strongest to weakest, which attains maximal in rearrangement inequality.
			\end{itemize}
		\end{remark}

		Tight bounds are generally unavailable when \gls{mimo} direct link is present.
		This is because the \gls{ris} needs to balance the backward-forward (multiplicative) and direct-indirect (additive) subspace alignments.

		% \begin{figure}[!t]
		% 	\centering
		% 	\resizebox{0.65\columnwidth}{!}{
		% 		\input{assets/simulation/pc_singular_pareto.tex}
		% 	}
		% 	\caption{Singular value Pareto frontier. $(N^\mathrm{T}, N^\mathrm{S}, N^\mathrm{R}) = (4, 64, 2)$, $(\Lambda^\mathrm{D}, \Lambda^\mathrm{F}, \Lambda^\mathrm{B}) = (0, -17.5, -17.5) \unit{dB}$.}
		% 	\label{sm:pc_singular_pareto}
		% \end{figure}

		% The Pareto frontier and evolving trend of channel singular values are shown in Fig.~\ref{sm:pc_singular_pareto} and \ref{sm:pc_singular_bound}.
		% Clearly, \gls{bd}-\gls{ris} with a larger group size can redistribute the channel singular values to a wider range.
	\end{subsection}

@@

\begin{section}{Rate Maximization}
	The \gls{mimo} rate maximization problem is formulated w.r.t. joint active and passive beamforming
	\begin{maxi!}
		{\scriptstyle{\mathbf{W},\mathbf{\Theta}}}{R = \log \det \biggl(\mathbf{I} + \frac{\mathbf{W}^\mathsf{H}\mathbf{H}^\mathsf{H}\mathbf{H}\mathbf{W}}{\eta}\biggr)}{\label{op:rate}}{\label{ob:rate}}
		\addConstraint{\lVert \mathbf{W} \rVert _\mathrm{F}^2 \le P}{}{}
		\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
	\end{maxi!}
	where $\mathbf{W}$ is the transmit precoder, $R$ is the achievable rate, $\eta$ is the noise power, and $P$ is the transmit power budget.
	Two methods are proposed below to solve problem \eqref{op:rate}.

>>>>

	\emph{Notation:}
	Italic, bold lower-case, and bold upper-case letters indicate scalars, vectors and matrices, respectively.
	$\jmath$ denotes the imaginary unit.
	$\mathbb{C}$ represents the set of complex numbers.
	$\mathbb{U}^{n \times n}$ denotes the set of $n \times n$ unitary matrices.
	$\mathbf{0}$ and $\mathbf{I}$ are the all-zero and identity matrices with appropriate size, respectively.
	$\Re\{\cdot\}$ takes the real part of a complex number.
	$\mathrm{tr}(\cdot)$ and $\det(\cdot)$ evaluates the trace and determinant of a square matrix, respectively.
	$\mathrm{diag}(\cdot)$ constructs a square matrix with arguments on the main diagonal and zeros elsewhere.
	$\mathrm{sv}(\cdot)$ returns the singular value vector.
	$\sigma_n(\cdot)$ and $\lambda_n(\cdot)$ is the $n$-th largest singular value and eigenvalue, respectively.
	% $\boldsymbol{\sigma}(\cdot)$ and $\boldsymbol{\lambda}(\cdot)$ are the corresponding vectors.
	$(\cdot)^*$, $(\cdot)^\mathsf{T}$, $(\cdot)^\mathsf{H}$, $(\cdot)^\dagger$ $(\cdot)^{(r)}$, $(\cdot)^{\star}$ denote the conjugate, transpose, conjugate transpose (Hermitian), Moore-Penrose inverse, $r$-th iterated point, and final solution, respectively.
	$(\cdot)_{[x:y]}$ is a shortcut for $(\cdot)_x,(\cdot)_{x+1},\ldots,(\cdot)_y$.
	$\lvert \cdot \rvert$ denotes the absolute value.
	% $\lVert \cdot \rVert _p$ means the $p$-norm and $\lVert \cdot \rVert$ suggests $p = 2$.
	$\lVert \cdot \rVert$ means the Euclidean norm.
	$\lVert \cdot \rVert _\mathrm{F}$ represents the Frobenius norm.
	$\mathcal{CN}(\mathbf{0}, \mathbf{\Sigma})$ is the multivariate \gls{cscg} distribution with mean $\mathbf{0}$ and covariance $\mathbf{\Sigma}$.
	$\sim$ means ``distributed as''.
\end{section}

@@

\begin{section}{Channel Singular Values Redistribution}
	\begin{subsection}{A Toy Example}
		We first illustrate the channel shaping capabilities of different \gls{ris} by a toy example.
		% We first use a toy example to illustrate that \gls{bd}-\gls{ris} can provide a wider dynamic range of channel singular values.
		Consider a $2 \times 2 \times 2$ setup where the direct link is blocked.
		The diagonal \gls{ris} is modeled by $\mathbf{\Theta}_\mathrm{D} = \mathrm{diag}(e^{\jmath \theta_1}, e^{\jmath \theta_2})$ while the unitary \gls{bd}-\gls{ris} has 4 independent angular parameters
		\begin{equation}
			\mathbf{\Theta}_\mathrm{U} = e^{\jmath \phi} \begin{bmatrix}
				e^{\jmath \alpha} \cos \psi  & e^{\jmath \beta} \sin \psi   \\
				-e^{-\jmath \beta} \sin \psi & e^{-\jmath \alpha} \cos \psi
			\end{bmatrix}.
		\end{equation}
		In particular, $\phi$ has no impact on the singular value because $\mathrm{sv}(e^{\jmath \phi} \mathbf{A}) = \mathrm{sv}(\mathbf{A})$.
		We also enforce symmetry by $\beta = \pi / 2$ such that both architectures have the same number of angular parameters.
		\begin{figure}
			\centering
			\includegraphics[width=\columnwidth]{assets/simulation/pc_singular_toy.eps}
			\caption{$2 \times 2 \times 2$ channel singular value shaping by diagonal and symmetry unitary \gls{ris}. Direct link is absent.}
			\label{sm:pc_singular_toy}
		\end{figure}
		Fig.~\ref{sm:pc_singular_toy} shows the channel singular values achieved by an exhaustive grid search over $(\theta_1, \theta_2)$ for diagonal \gls{ris} and $(\alpha, \psi)$ for symmetric unitary \gls{ris}.
		It is observed that both singular values can be manipulated up to $9\%$ using diagonal \gls{ris} and $42\%$ using symmetric \gls{bd}-\gls{ris}, despite both architectures have the same number of scattering elements and design parameters.
		A larger performance gap is expected when asymmetric \gls{bd}-\gls{ris} is available.
		This example shows \gls{bd}-\gls{ris} can provide a wider dynamic range of channel singular values and motivates further studies on channel shaping.
	\end{subsection}

@@

		% The Pareto singular value problem \eqref{op:pareto} can thus be solved by Algorithm~\ref{ag:rcg}.
		\begin{lemma}\label{lm:pareto_gradient}
			The Euclidean gradient of \eqref{ob:pareto} w.r.t. \gls{bd}-\gls{ris} group $g$ is
			\begin{equation}
				\frac{\partial \sum_n \rho_n \sigma_n(\mathbf{H})}{\partial \mathbf{\Theta}_g^*} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \mathbf{U} \mathrm{diag}(\rho_1,\ldots,\rho_N) \mathbf{V}^\mathsf{H} \mathbf{H}_{\mathrm{F},g}^\mathsf{H},
				\label{eq:pareto_gradient}
			\end{equation}
			where $\mathbf{U} \in \mathbb{C}^{N_\mathrm{R} \times N}$ and $\mathbf{V} \in \mathbb{C}^{N_\mathrm{T} \times N}$ are the left and right compact singular matrices of $\mathbf{H}$, respectively.
		\end{lemma}
		\begin{proof}
			Let $\mathbf{H} = \sum_n \mathbf{u}_n \sigma_n \mathbf{v}_n^\mathsf{H}$ be the compact \gls{svd} of the equivalent channel.
			Since the singular vectors are orthonormal, the $n$-th singular value can be expressed as
			\begin{equation}
				\sigma_n = \mathbf{u}_n^\mathsf{H} \mathbf{H} \mathbf{v}_n = \mathbf{u}_n^\mathsf{T} \mathbf{H}^* \mathbf{v}_n^*,
			\end{equation}
			whose differential w.r.t. $\mathbf{\Theta}_g^*$ is
			\begin{align*}
				\partial \sigma_n
				& = \partial \mathbf{u}_n^\mathsf{T} \underbrace{\mathbf{H}^* \mathbf{v}_n^*}_{\sum_m \mathbf{u}_m^* \sigma_m \mathbf{v}_m^\mathsf{T} \mathbf{v}_n} + \mathbf{u}_n^\mathsf{T} \cdot \partial \mathbf{H}^* \cdot \mathbf{v}_n^* + \underbrace{\mathbf{u}_n^\mathsf{T} \mathbf{H}^*}_{\mathbf{u}_n^\mathsf{T} \sum_m \mathbf{u}_m^* \sigma_m \mathbf{v}_m^\mathsf{T}} \partial \mathbf{v}_n^*\\
				& = \underbrace{\partial \mathbf{u}_n^\mathsf{T} \mathbf{u}_n^*}_{\partial 1 = 0} \cdot \sigma_n + \mathbf{u}_n^\mathsf{T} \cdot \partial \mathbf{H}^* \cdot \mathbf{v}_n^* + \sigma_n \cdot \underbrace{\mathbf{v}_n^\mathsf{T} \partial \mathbf{v}_n^*}_{\partial 1 = 0}\\
				& = \mathbf{u}_n^\mathsf{T} \mathbf{H}_{\mathrm{B},g}^* \cdot \partial \mathbf{\Theta}_g^* \cdot \mathbf{H}_{\mathrm{F},g}^* \mathbf{v}_n^*\\
				& = \mathrm{tr}(\mathbf{H}_{\mathrm{F},g}^* \mathbf{v}_n^*\mathbf{u}_n^\mathsf{T} \mathbf{H}_{\mathrm{B},g}^* \cdot \partial \mathbf{\Theta}_g^*).
			\end{align*}
			According to \cite{Hjorungnes2007}, the corresponding complex derivative is
			\begin{equation}
				\frac{\partial \sigma_n}{\partial \mathbf{\Theta}_g^*} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \mathbf{u}_n \mathbf{v}_n^\mathsf{H} \mathbf{H}_{\mathrm{F},g}^\mathsf{H}.
				\label{eq:derivative_sv}
			\end{equation}
			Linear combination of \eqref{eq:derivative_sv} yields \eqref{eq:pareto_gradient}.
		\end{proof}

@@

	\begin{subsection}{Some Analytical Bounds}\label{ss:bounds}
		We then discuss some analytical bounds related to channel singular values.
		\begin{proposition}[rank-deficient channel]\label{pp:dof}
			In point-to-point \gls{mimo}, \gls{bd}-\gls{ris} cannot achieve a higher \gls{dof} than diagonal \gls{ris}.
		\end{proposition}
		\begin{proof}
			The scattering matrix of \gls{bd}-\gls{ris} can be decomposed as\footnote{This is because (block) unitary matrices are closed under multiplication.}
			\begin{equation}
				\mathbf{\Theta} = \mathbf{L} \mathbf{\Theta}_\mathrm{D} \mathbf{R}^\mathsf{H},
			\end{equation}
			where $\mathbf{\Theta}_\mathrm{D} \in \mathbb{U}^{N_\mathrm{S} \times N_\mathrm{S}}$ corresponds to diagonal \gls{ris} and $\mathbf{L}, \mathbf{R} \in \mathbb{U}^{N_\mathrm{S} \times N_\mathrm{S}}$ are block diagonal matrices of $L \times L$ unitary blocks.
			Manipulating $\mathbf{L}$ and $\mathbf{R}$ rotates the linear spans of $\bar{\mathbf{H}}_\mathrm{B} \triangleq \mathbf{H}_\mathrm{B} \mathbf{L}$ and $\bar{\mathbf{H}}_\mathrm{F} \triangleq \mathbf{R}^\mathsf{H} \mathbf{H}_\mathrm{F}$ and maintains their rank.
			On the other hand, there exists a $\mathbf{\Theta}_\mathrm{D}$ such that
			\begin{equation*}
				\begin{split}
					\mathrm{rank}(\mathbf{H}_\mathrm{B} \mathbf{\Theta}_\mathrm{D} \mathbf{H}_\mathrm{F})
					& = \min \bigl( \mathrm{rank}(\mathbf{H}_\mathrm{B}), \mathrm{rank}(\mathbf{\Theta}_\mathrm{D}), \mathrm{rank}(\mathbf{H}_\mathrm{F}) \bigr) \\
					& = \min \bigl( \mathrm{rank}(\bar{\mathbf{H}}_\mathrm{B}), N_\mathrm{S}, \mathrm{rank}(\bar{\mathbf{H}}_\mathrm{F}) \bigr) \\
					& = \max_\mathbf{\Theta} \ \mathrm{rank}(\mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F})
				\end{split}
			\end{equation*}
			The same result holds if the direct link is present.
		\end{proof}

		\begin{proposition}[\gls{los} forward\footnote{A similar result holds for \gls{los} backward channel.} channel]\label{pp:los}
			If the forward channel is rank-1, then \gls{bd}-\gls{ris} can at most enlarge (resp. suppress) the $n$-th ($n \ge 2$) channel singular value to the $(n-1)$-th (resp. $n$-th) singular value of $\mathbf{T}$, that is,
			\begin{equation}
				\sigma_1(\mathbf{T}) \ge {\sigma_2(\mathbf{H})} \ge \sigma_2(\mathbf{T}) \ge \ldots \ge \sigma_{N-1}(\mathbf{T}) \ge {\sigma_N(\mathbf{H})} \ge \sigma_N(\mathbf{T}),
				\label{iq:sv_bound_los}
			\end{equation}
			where $\mathbf{T} \mathbf{T}^\mathsf{H} = \mathbf{H}_\mathrm{D} (\mathbf{I} - \mathbf{v}_\mathrm{F} \mathbf{v}_\mathrm{F}^\mathsf{H}) \mathbf{H}_\mathrm{D}$ and $\mathbf{v}_\mathrm{F}$ is the right compact singular vector of $\mathbf{H}_\mathrm{F}$.
			Note that $\sigma_1(\mathbf{H})$ is unbounded with a sufficiently large \gls{ris}.
		\end{proposition}
		\begin{proof}
			Let $\mathbf{H}_\mathrm{F} = \sigma_\mathrm{F} \mathbf{u}_\mathrm{F} \mathbf{v}_\mathrm{F}^\mathsf{H}$ be the compact \gls{svd} of the forward channel.
			The channel Gram matrix can be written as Hermitian-plus-rank-1:
			\begin{equation}
				\mathbf{G} \triangleq \mathbf{H} \mathbf{H}^\mathsf{H} = \mathbf{Y} + \mathbf{z} \mathbf{z}^\mathsf{H},
			\end{equation}
			where $\mathbf{Y} \triangleq \mathbf{H}_\mathrm{D}^\mathsf{H} (\mathbf{I} - \mathbf{v}_\mathrm{F} \mathbf{v}_\mathrm{F}^\mathsf{H}) \mathbf{H}_\mathrm{D}$ and $\mathbf{z} \triangleq \sigma_\mathrm{F} \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{u}_\mathrm{F} + \mathbf{H}_\mathrm{D} \mathbf{v}_\mathrm{F}$.
			By the Cauchy interlacing formula \cite{Golub2013}, the $n$-th ($n \ge 2$) eigenvalues of $\mathbf{G}$ are bounded by
			\begin{equation}
				\lambda_1(\mathbf{Y}) \ge \lambda_2(\mathbf{G}) \ge \lambda_2(\mathbf{Y}) \ge \ldots \ge \lambda_{N-1}(\mathbf{Y}) \ge \lambda_N(\mathbf{G}) \ge \lambda_N(\mathbf{Y}).
				\label{iq:ev_bound_los}
			\end{equation}
			Since $\mathbf{Y} = \mathbf{T} \mathbf{T}^\mathsf{H}$ is positive semi-definite, the eigenvalues are non-negative and taking the square root of \eqref{iq:ev_bound_los} gives \eqref{iq:sv_bound_los}.
		\end{proof}
		It is worth notice that a similar conclusion holds for diagonal \gls{ris} \cite{Semmler2023}.
		We will later show that for a finite $N_\mathrm{S}$, using a larger group size can approach those bounds better.

@@

		\begin{remark}
			Interestingly, \eqref{iq:sv_bound_fc}--\eqref{iq:sv_bound_fc_power} are simultaneously tight when $\mathbf{\Theta} = \mathbf{V}_\mathrm{B} \mathbf{U}_\mathrm{F}^\mathsf{H}$.
			From \eqref{eq:scattering_fc} and \eqref{eq:channel_equivalent_fc}, we conclude the off-diagonal entries can enhance the capabilities of
			\begin{itemize}
				\item subspace alignment: $\mathbf{V}_\mathrm{B}$ and $\mathbf{U}_\mathrm{F}^\mathsf{H}$ in \eqref{eq:scattering_fc} fully align the subspaces of $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ by rotation;
				\item subchannel rearrangement: $\mathbf{X} = \mathbf{I}$ in \eqref{eq:channel_equivalent_fc} pairs the subchannels of $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ from strongest to weakest, which attains the maximal in rearrangement inequality.
			\end{itemize}
		\end{remark}

		Tight bounds are generally unavailable when \gls{mimo} direct link is present.
		This is because the \gls{ris} needs to balance the direct-indirect (additive) and backward-forward (multiplicative) subspace alignments.
	\end{subsection}

@@

\begin{section}{Achievable Rate Maximization}
	The \gls{mimo} achievable rate maximization problem is formulated w.r.t. joint active and passive beamforming
	\begin{maxi!}
		{\scriptstyle{\mathbf{W},\mathbf{\Theta}}}{R = \log \det \biggl(\mathbf{I} + \frac{\mathbf{W}^\mathsf{H}\mathbf{H}^\mathsf{H}\mathbf{H}\mathbf{W}}{\eta}\biggr)}{\label{op:rate}}{\label{ob:rate}}
		\addConstraint{\lVert \mathbf{W} \rVert _\mathrm{F}^2 \le P}{}{}
		\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
	\end{maxi!}
	where $\mathbf{W}$ is the transmit precoder, $R$ is the achievable rate, $\eta$ is the noise power, and $P$ is the transmit power budget.
	Two methods are proposed below to solve problem \eqref{op:rate}.

	\begin{subsection}{Alternating Optimization}
		Consider an \gls{ao} approach that updates $\mathbf{\Theta}$ and $\mathbf{W}$ iteratively.
		For a given $\mathbf{W}$, the passive beamforming subproblem is
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\log \det \biggl(\mathbf{I} + \frac{\mathbf{H} \mathbf{Q} \mathbf{H}^\mathsf{H}}{\eta}\biggr)}{\label{op:rate_passive}}{\label{ob:rate_passive}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
		\end{maxi!}
		where $\mathbf{Q} \triangleq \mathbf{W} \mathbf{W}^\mathsf{H}$ is the transmit covariance matrix.
		\begin{lemma}\label{lm:rate_gradient}
			The Euclidean gradient of \eqref{ob:rate_passive} w.r.t. \gls{bd}-\gls{ris} block $g$ is
			\begin{equation}
				\frac{\partial R}{\partial \mathbf{\Theta}_g^*} = \frac{1}{\eta} \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \biggl(\mathbf{I} + \frac{\mathbf{H}\mathbf{Q}\mathbf{H}^\mathsf{H}}{\eta}\biggr)^{-1} \mathbf{H} \mathbf{Q} \mathbf{H}_{\mathrm{F},g}^\mathsf{H}.
				\label{eq:rate_gradient}
			\end{equation}
		\end{lemma}
